---
title: "Scoring questionnaires"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring questionnaires}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(colleyRstats)
```

Turning raw item columns into subscale scores is the most repeated step of a
user study, and the one most able to be wrong without anything looking wrong.
Every project re-implements the same reverse-coding and averaging by hand, and a
single mis-numbered item changes every result downstream without producing a
single error.

`score_questionnaire()` applies a published instrument's own scoring key.

## What is available

```{r}
list_questionnaires()[, c("key", "name", "n_items", "n_subscales", "scale")]
```

## Scoring

Point it at the item columns. A prefix is usually enough:

```{r}
set.seed(1)
study <- data.frame(participant = factor(1:8))
study[paste0("sus_", 1:10)] <- lapply(1:10, function(i) sample(1:5, 8, TRUE))

score_questionnaire(study, "sus", prefix = "sus_")
```

The SUS score is the sum of the ten recoded items times 2.5, which puts it on
0--100. That is a percentage of the maximum, **not** a percentile: the mean SUS
across published studies is about 68, so 68 is average rather than poor.

A prefix selects the columns, and their *names* then say which item each one
holds -- never their sort order, which would put `sus_10` before `sus_2`, or
`tlx_effort` before `tlx_mental`. Either what follows the prefix is the item's
code or label, compared without case or punctuation:

```{r}
tlx_named <- data.frame(
  tlx_effort = 50, tlx_frustration = 60, tlx_mental = 10,
  tlx_performance = 40, tlx_physical = 20, tlx_temporal = 30
)
score_questionnaire(tlx_named, "nasa_tlx", prefix = "tlx_")$Mental_Demand
```

or it carries one item number per column, running exactly 1 to the number of
items (`sus_2`, `SUS[2]`, or `SUS_2_1`, where the number that varies across the
columns is the item). Names that do neither -- a partial match, numbering from
0 -- are an error asking for a named `items` mapping rather than a guess.

### Item columns must be numbers

Survey exports often deliver responses as text. A column holding
`"Strongly agree"`, `"5 - Strongly agree"` or a decimal comma such as `"2,5"`
stops scoring with an error that names the values and says how to recode them:
converting them would silently turn those answers into missing values, and with
a relaxed `min_valid` the row would be scored as if the participant had skipped
them. Blank cells are missing responses. Factors are read by their labels when
those are numbers, and by level order only when the factor is ordered.

```{r}
text_export <- study
text_export$sus_3 <- as.character(text_export$sus_3)
text_export$sus_3[1] <- "Strongly agree"
try(score_questionnaire(text_export, "sus", prefix = "sus_"))
```

Every instrument's notes say this sort of thing, and
`questionnaire_items()` prints them:

```{r}
questionnaire_items("ueq_s")
```

### Declaring the response range

Instruments are administered on whatever scale the survey tool offered. Say
which one you used and the responses are rescaled onto the instrument's own
range before scoring:

```{r}
tlx <- data.frame(
  mental = c(14, 3), physical = c(2, 1), temporal = c(11, 4),
  performance = c(6, 2), effort = c(13, 5), frustration = c(9, 1)
)

# The 21-point NASA-TLX sheet, scored onto the conventional 0-100
score_questionnaire(tlx, "nasa_tlx", scale = c(1, 21))
```

Responses outside the range you declare are an error rather than a silent
rescale, which is what stops a 1--7 export from being scored as if it were the
0--100 TLX.

The quieter failure is a coding that fits the range but is shifted against it.
The scoring message prints the observed range next to the assumed one, and
without an explicit `scale` three patterns draw a warning: NASA-TLX responses
that never exceed 21 (the 21-point sheet, which scored as 0--100 understates
workload about five-fold); no response at 0 on a zero-based instrument such as
the IPQ or SSQ (a 1-based export in which nobody chose the top point); and no
response at the top of a one-based instrument such as the SUS (a 0-based export
in which nobody chose 0). The last two need at least five respondents. Passing
`scale` explicitly confirms the coding and silences them.

```{r}
# The same 21-point responses, without saying so: RTLX comes out five times too low
score_questionnaire(tlx, "nasa_tlx")$RTLX
```

A fractional response on an instrument answered in whole scale points (every
built-in one except the NASA-TLX) warns too: it is an averaged, imputed or
mis-exported column rather than a response.

### Adding scores to the data

```{r}
scored <- score_questionnaire(study, "sus", prefix = "sus_", append = TRUE)
names(scored)[1:3]
scored$SUS
```

## Verify the mapping before you trust the scores

This is the important part.

Item numbers, item order and item polarity belong to the sheet a study actually
administered, not to the instrument in the abstract. Survey tools renumber
items, translations reorder them, short forms drop them from the middle, and
semantic differentials get printed with the poles the other way round. This
package applies each instrument's *published* key — the right default, and still
only a default. Whatever your columns are called, the instrument's item 3 is
taken to be the question your participants saw third; if your sheet ordered,
worded or poled them differently, the scores look entirely reasonable and are
wrong. (An unnamed `items` vector is mapped by position, too.)

AttrakDiff is the case in point. The `attrakdiff` key lists its 28 word pairs
blocked by dimension, each with its negative term first -- the layout of data
that has already been re-sorted and re-poled, not of the official form. Data
stored as answered on the official sheet, which interleaves the pairs and
prints 15 of them with the positive term on the left, belong to
`"attrakdiff_official"`. Both keys use the same item codes for the same word
pairs, so a named `items` mapping works with either.

Nothing errors when that happens, which is why R says it out loud: the first
time each distinct mapping is scored in a session, the mapping and a caution
print together. Read them rather than tuning them out, and double-check any
number before it reaches a paper.

Run this once per instrument, per study, and read what it prints:

```{r}
invisible(check_questionnaire(study, "sus", prefix = "sus_"))
```

It shows which column supplies which item, which subscale that item loads on,
whether it is reverse-coded, and the observed range of each column -- so a
column of all-3s where you expected variation, or a 1--7 export where the
instrument expects 1--5, is visible immediately.

If your survey printed a pair the other way round, `reverse_items` reverses it
in addition to the key's own reversals:

```{r}
# A survey that anchored NASA-TLX performance "Good" at the high end
scores <- score_questionnaire(tlx, "nasa_tlx",
  scale = c(1, 21),
  reverse_items = "performance"
)
scores$Performance
```

Do not pass an instrument's own reverse set there. `reverse_items` *toggles*,
for backward compatibility, so `reverse_items = c(2, 4, 6, 8, 10)` on the SUS
-- a line many tutorials contain -- un-reverses all five negatively worded
items, and a perfect respondent scores 50. That now raises a warning. If an
export genuinely stored items that the key reverses already reversed, say so
with `unreverse_items`, which does the same on purpose and without a warning:

```{r}
perfect <- data.frame(t(c(5, 1, 5, 1, 5, 1, 5, 1, 5, 1)))
names(perfect) <- paste0("sus_", 1:10)
score_questionnaire(perfect, "sus", prefix = "sus_")$SUS

# Items 2 and 4 were stored pre-reversed by the survey tool
pre_reversed <- perfect
pre_reversed[c("sus_2", "sus_4")] <- 5
score_questionnaire(pre_reversed, "sus", prefix = "sus_",
  unreverse_items = c("sus2", "sus4")
)$SUS
```

A named `items` argument avoids positional mapping altogether, and is the safer
choice for an export you did not lay out yourself:

```{r}
score_questionnaire(
  tlx, "nasa_tlx",
  scale = c(1, 21),
  items = c(
    mental = "mental", physical = "physical", temporal = "temporal",
    performance = "performance", effort = "effort", frustration = "frustration"
  )
)$RTLX
```

## Incomplete responses

By default a subscale is scored only if every one of its items was answered,
and is `NA` otherwise. Nothing is imputed silently.

```{r}
gappy <- study
gappy$sus_3[1] <- NA

score_questionnaire(gappy, "sus", prefix = "sus_")$SUS[1]

# Score responses that are at least 80% complete
score_questionnaire(gappy, "sus", prefix = "sus_", min_valid = 0.8)$SUS[1]
```

With `min_valid` relaxed, the missing items are completed by the instrument's
published rule. For the SUS that is Brooke's (1996): a respondent who cannot
answer an item marks the centre point, so each missing item counts as 3 on
1--5. One missing item in an otherwise perfect response gives 95, not 100. For
every other instrument a subscale is the mean of the items present, with a
summed score scaled up proportionally so it stays on its published range.
`impute = "prorate"` or `impute = "midpoint"` chooses a rule explicitly --
report which one you used:

```{r}
score_questionnaire(gappy, "sus", prefix = "sus_", min_valid = 0.8,
  impute = "prorate"
)$SUS[1]
```

## Internal consistency

`score_reliability()` computes Cronbach's alpha on the same recoded matrix that
is aggregated, so reverse-coded items are already flipped. It reports every
subscale, plus the whole scale where the instrument defines an overall score --
the ten-item SUS alpha that SUS papers report:

```{r}
score_reliability(study, "sus", prefix = "sus_")
```

Because the reversal has already happened, a negative alpha means a real
problem -- the items of that subscale do not point the same way -- rather than a
forgotten flip. It warns when one appears.

McDonald's omega (total) comes from an unrotated one-factor solution
(`psych::fa()`), so it needs 'psych' but not 'GPArotation', and -- unlike
`psych::omega()` -- does not flip negatively loading items, which would hide
the very reversal error a negative alpha exposes. For two-item scales, such as
SUS Learnability, omega is not identified and alpha understates reliability
whenever the two item variances differ; `spearman_brown` gives the coefficient
Eisinga, te Grotenhuis and Pelzer (2013) recommend instead. The `note` column
says why any coefficient is missing.

Do not read the values above as anything: the example responses are random, so
any coefficient that looks respectable does so purely by chance. Eight
participants is far too few to estimate reliability at all, which is itself
worth remembering -- `n_complete` is reported next to it for that reason.

## Your own instruments

`define_questionnaire()` registers one, after which it behaves exactly like a
built-in. Put the call in a project's setup script and every analysis in that
project scores it identically:

```{r}
define_questionnaire(
  key = "acceptance",
  name = "Van der Laan acceptance scale",
  reference = "Van der Laan, Heino & De Waard (1997), Transp. Res. C 5(1)",
  scale = c(-2, 2),
  label = c(
    "useful - useless", "pleasant - unpleasant", "bad - good",
    "nice - annoying", "effective - superfluous", "irritating - likeable",
    "assisting - worthless", "undesirable - desirable",
    "raising alertness - sleep-inducing"
  ),
  subscale = rep(c("Usefulness", "Satisfying"), length.out = 9),
  reverse = c(1, 2, 4, 5, 7, 9),
  higher = "better"
)

vdl <- as.data.frame(matrix(c(-2, 2, -2, 2, -2, 2, -2, 2, -2), nrow = 1))
names(vdl) <- paste0("item", 1:9)

score_questionnaire(vdl, "acceptance")
```

## Sickness over time

FMS and MISC are single items sampled repeatedly during exposure, so the
analysis lives in the time course rather than in any one measurement.
`summarize_sickness()` produces the per-participant measures those studies
report:

```{r}
ratings <- data.frame(
  participant = rep(c("p1", "p2"), each = 5),
  minute = rep(0:4, 2),
  fms = c(0, 1, 3, 6, 8, 0, 0, 1, 1, 2)
)

summarize_sickness(ratings,
  value = "fms", id = "participant",
  time = "minute", threshold = 5
)
```

`auc` is the trapezoidal area under the rating curve and `auc_rate` is that
divided by the observed duration, i.e. the time-weighted mean rating -- which is
comparable across participants who were exposed for different lengths of time.

Note that MISC is *ordinal* and unevenly spaced: the step from 6 (nausea) to 10
(vomiting) is not four times the step from 0 to 1. Analyse it with an ordinal
model rather than by taking means.

You have to say so, though. A 0--10 MISC column has more distinct values than
`classify_outcome()`'s `ordinal_max_levels` of 7, so left to itself it is taken
for a count and `recommend_test()` proposes a Poisson GLMM:

```{r}
misc <- data.frame(misc = c(0, 1, 3, 6, 10, 2, 4, 8))
classify_outcome(score_questionnaire(misc, "misc")$MISC)
```

Pass `outcome_type = "ordinal"` to `fit_recommended()`, or raise
`ordinal_max_levels` to 11.

## Onward

`fit_recommended()` takes the scored data and fits the model each outcome calls
for. Here is the whole path, from raw items to a manuscript sentence:

```{r}
set.seed(7)
trial <- expand.grid(
  participant = factor(1:12),
  condition = factor(c("baseline", "ambient", "explicit"))
)
effect <- c(baseline = 0, ambient = 0.6, explicit = 1.2)[trial$condition]
# Participants differ from one another too, which is what the random intercept
# of the mixed model is there to absorb.
person <- rnorm(12, 0, 0.6)[as.integer(trial$participant)]
for (i in 1:10) {
  raw <- 3 + effect + person + rnorm(nrow(trial), 0, 0.5)
  if (i %% 2 == 0) raw <- 6 - raw   # the even SUS items are negatively worded
  trial[[paste0("sus_", i)]] <- pmin(pmax(round(raw), 1), 5)
}

scored <- score_questionnaire(trial, "sus", prefix = "sus_", append = TRUE)

fit <- fit_recommended(
  scored,
  outcome = "SUS",
  predictors = "condition",
  cluster = "participant",
  verbose = FALSE
)

fit$recommendation$recommendation
fit$text
```

The sentence is LaTeX; `expand_latex_macros()` renders it as plain text if you
want to read it here rather than paste it into a manuscript.

Watch the classification when an outcome stays whole-numbered. A raw NASA-TLX
subscale scored onto 0--100 moves in steps of 5 and is recognised as a bounded
score (continuous), but a MISC rating or another whole-numbered rating is taken
for a count -- with a message saying so -- unless you pass
`outcome_type = "continuous"` or `"ordinal"`. SUS and RTLX are fractional, so
they classify as continuous on their own.

`use_study_project()` scaffolds a whole analysis around this -- scoring, models,
figures, and the generated LaTeX a manuscript reads -- as a `targets` pipeline,
and writes those `outcome_type` declarations for you.
